---
title: The Monopolists Are Winning
output:
    html_document:
        toc: true
        toc_float: true
        code_folding: hide
---

```{r include=FALSE}
library('knitr')

opts_chunk$set(fig.width = 9, fig.height = 9, fig.align = 'center',
               warning=FALSE)
```

# Concentration Is Growing in The United States
In recent years, the United States has seen a wave of big mergers and
acquisitions. Few of these deals have been challenged by the Federal Trade
Commision. As a result, many sectors of the economy have grown more
concentrated---the top firms in each sector have increased their market share.

```{r include=FALSE}
source('plots/concentration-growth.r')
read_chunk('plots/concentration-growth.r')
```
```{r plot.concentration}
```

The US Census Bureau measures concentration by calculating the percentage of
total revenue within an industry captured by the top firms ("top" being
determined by absolute firm revenue). The Census Bureau publishes four
different concentration percentages for each industry: one for the top four
firms, one for the top eight, one for the top 20, and one for the top 50.

We can use these measures to explore how each "class" of firms has increased
its share of total industry revenue over time. A firm is either one of the top
four firms, one of the next four firms, one of the next 12 firms after that, or
one of the remaining 30 firms in the top 50.

Between 2007 and 2012, the median industry saw concentration increase among all
of these firm classes except the "Next Four Firms" class (where it held
steady). Since concentration did not decrease in any of these classes, in the
median industry, the top 50 firms increased their share of revenue overall.

What is especially interesting here is the variability of concentration change
in the "Top Four Firms" class. It looks like in some industries the top four
"win," while in other industries the top four "lose." While concentration
nudged upward in most industries, this suggests that major shifts in
concentration may be due primarily to whether or not the top four firms were
able to consolidate their position.

```{r include=FALSE}
source('plots/concenfi.r')
read_chunk('plots/concenfi.r')
```
```{r plot.concenfi}
```

# Information and Media Are the Most Concentrated
The US Census Bureau categorizes firms according to the North American Industry
Classification System (NAICS), which assigns each firm an industry, industry
group, subsector, and sector. Of all the NAICS sectors, the Information sector,
which includes technology and media companies, has become the most concentrated
sector. The Utilities sector and the Finance and Insurance sector are also
highly concentrated.

```{r include=FALSE}
source('plots/sectors.r')
read_chunk('plots/sectors.r')
```
```{r plot.sectors}
```

# Business Formation Is Falling
As industries grow more concentrated, the major firms can use their market
power to erect barriers to entry. It can become more difficult for existing
firms to enter the market or for new firms to form in the first place. As
industries concentrate in the United States, we would therefore expect it to
become harder to form new businesses.

The US Census Bureau also tracks the formation of new businesses. Business
formation has indeed been falling in every state, even those states known as
hotbeds for startup technology ventures.

```{r include=FALSE}
source('plots/business-formation.r')
read_chunk('plots/business-formation.r')
```
```{r plot.formation}
```

More new business formation happens in large, populous states than in small
states, obviously. But if we look at per capita business formation, we can see
that western states, especially those western states not on the coast, have
historically had what we might call a high rate of entrepreneuship. But these
states have seen the most drastic declines in per capita business formation
over the decade between 2005 and 2015.

```{r include=FALSE}
source('plots/entrepreneurship.r')
read_chunk('plots/entrepreneurship.r')
```
```{r plot.entrepreneurship}
```

The geographic pattern here is particularly easy to see on a map:
```{r include=FALSE}
source('plots/map-entrepreneurship.r')
read_chunk('plots/map-entrepreneurship.r')
```
```{r plot.map.entrepreneurship}
```

# Consumer Complaints Are Rising
The Consumer Financial Protection Bureau (CFPB) was created in the wake of the
2008 financial crisis. The CFPB is meant to provide information to consumers in
the financial sector and address complaints that consumers have about specific
financial companies.

The CFPB keeps a record of every complaint the agency receives. These
complaints are specific only to the financial sector. But they provide some
measure of the market power that financial firms have over consumers---the
fewer competitors a firm has, the more it can afford to adopt consumer-hostile
practices.

Complaints have been rising in every state of the nation since the CFPB was
created in 2011.

```{r include=FALSE}
source('plots/sa-complaints.r')
read_chunk('plots/sa-complaints.r')
```
```{r plot.complaints}
```

The CFPB categorizes the complaints it receives according to a product and a
subproduct. The vast majority of complaints received by the CFPB are against
debt collection agencies or mortage providers. Other major sources of
complaints are credit score companies and student loan companies.

```{r include=FALSE}
source('plots/products.r')
read_chunk('plots/products.r')
```
```{r plot.products}
```
